---
title: "ComplexHeatmap"
author: "Brandon"
date: "12/23/2021"
output: 
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#https://bioconductor.org/packages/release/bioc/manuals/ComplexHeatmap/man/ComplexHeatmap.pdf
#https://github.com/jokergoo/ComplexHeatmap
#https://jokergoo.github.io/ComplexHeatmap-reference/book/oncoprint.html#oncoprint-general-settings

```

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

#BiocManager::install("ComplexHeatmap")

library(ComplexHeatmap)
```
# Chapter 1 Making a Simple Heatmap 
## Chpater 1.1 Creating the Dataset
```{r}
# Making the dataset 

mat <- matrix(rnorm(18*24),nrow = 18,ncol=24)

#Something more specific
set.seed(123)
nr1 = 4; nr2 = 8; nr3 = 6; nr = nr1 + nr2 + nr3
nc1 = 6; nc2 = 8; nc3 = 10; nc = nc1 + nc2 + nc3
mat = cbind(rbind(matrix(rnorm(nr1*nc1, mean = 1,   sd = 0.5), nr = nr1),
          matrix(rnorm(nr2*nc1, mean = 0,   sd = 0.5), nr = nr2),
          matrix(rnorm(nr3*nc1, mean = 0,   sd = 0.5), nr = nr3)),
    rbind(matrix(rnorm(nr1*nc2, mean = 0,   sd = 0.5), nr = nr1),
          matrix(rnorm(nr2*nc2, mean = 1,   sd = 0.5), nr = nr2),
          matrix(rnorm(nr3*nc2, mean = 0,   sd = 0.5), nr = nr3)),
    rbind(matrix(rnorm(nr1*nc3, mean = 0.5, sd = 0.5), nr = nr1),
          matrix(rnorm(nr2*nc3, mean = 0.5, sd = 0.5), nr = nr2),
          matrix(rnorm(nr3*nc3, mean = 1,   sd = 0.5), nr = nr3))
   )
mat = mat[sample(nr, nr), sample(nc, nc)] # random shuffle rows and columns
rownames(mat) = paste0("row", seq_len(nr))
colnames(mat) = paste0("column", seq_len(nc))

Heatmap(mat)

```
## Chapter 1.2 Changing the Colours  

```{r}
library(circlize)
col_fun = colorRamp2(c(-2, 0, 2), c("green", "white", "red"))
col_fun(seq(-3, 3))

Heatmap(mat, name = "mat", col = col_fun)
```


```{r}
mat2 = mat
mat2[1, 1] = 100000
Heatmap(mat2, name = "mat", col = col_fun, 
    column_title = "a matrix with outliers")
```


```{r}
Heatmap(mat, name = "mat", col = col_fun, column_title = "mat")
Heatmap(mat/4, name = "mat", col = col_fun, column_title = "mat/4")
Heatmap(abs(mat), name = "mat", col = col_fun, column_title = "abs(mat)")
```


```{r}
Heatmap(mat, name = "mat", col = rev(rainbow(10)), 
    column_title = "set a color vector for a continuous matrix")

discrete_mat = matrix(sample(1:4, 100, replace = TRUE), 10, 10)
colors = structure(1:4, names = c("1", "2", "3", "4")) # black, red, green, blue
Heatmap(discrete_mat, name = "mat", col = colors,
    column_title = "a discrete numeric matrix")

discrete_mat = matrix(sample(letters[1:4], 100, replace = TRUE), 10, 10)
colors = structure(1:4, names = letters[1:4])
Heatmap(discrete_mat, name = "mat", col = colors,
    column_title = "a discrete character matrix")
```
```{r}
mat_with_na = mat
na_index = sample(c(TRUE, FALSE), nrow(mat)*ncol(mat), replace = TRUE, prob = c(1, 9))
mat_with_na[na_index] = NA
Heatmap(mat_with_na, name = "mat", na_col = "black",
    column_title = "a matrix with NA values")
```


## Chapter 1.2 Changing the tittles

```{r}
Heatmap(mat, 
        name = "mat", 
        column_title = "I am a column title", 
        row_title = "I am a row title")
```
```{r}
Heatmap(mat, name = "mat", 
        column_title = "I am a column title at the bottom", 
        column_title_side = "bottom") # location of the tittles

Heatmap(mat, name = "mat", 
        column_title = "I am a big column title", 
        column_title_gp = gpar(fontsize = 20, fontface = "bold"))

Heatmap(mat, name = "mat", 
        column_title = expression(hat(beta) == (X^t * X)^{-1} * X^t * y)) 
```
## Chapter 1.3 Clustering the Heatmap   

```{r}
Heatmap(mat, name = "mat", cluster_rows = FALSE,cluster_columns = FALSE) # turn off clustering
Heatmap(mat, name = "mat", cluster_rows = TRUE, cluster_columns = TRUE) # turn on clustering
```



```{r}
Heatmap(mat, name = "mat", cluster_rows = TRUE, cluster_columns = TRUE) # turn on clustering
Heatmap(mat, name = "mat", show_column_dend = FALSE, show_row_dend = FALSE) # hide column dendrogram
```

```{r}
Heatmap(mat, name = "mat", column_dend_height = unit(2, "cm"), 
    row_dend_width = unit(2, "cm"))

Heatmap(mat, name = "mat", column_dend_height = unit(4, "cm"), 
    row_dend_width = unit(4, "cm"))
```
## Chapter 1.3.1 Distance methods

```{r}
Heatmap(mat, name = "mat", clustering_distance_rows = "pearson",
    column_title = "pre-defined distance method (1 - pearson)")

Heatmap(mat, name = "mat", clustering_distance_rows = function(m) dist(m),
    column_title = "a function that calculates distance matrix")

Heatmap(mat, name = "mat", clustering_distance_rows = function(x, y) 1 - cor(x, y),
    column_title = "a function that calculates pairwise distance")

```

```{r}
mat_with_outliers = mat
for(i in  1:10) mat_with_outliers[i, i] = 1000
robust_dist = function(x, y) {
    qx = quantile(x, c(0.1, 0.9))
    qy = quantile(y, c(0.1, 0.9))
    l = x > qx[1] & x < qx[2] & y > qy[1] & y < qy[2]
    x = x[l]
    y = y[l]
    sqrt(sum((x - y)^2))
}

Heatmap(mat_with_outliers, name = "mat", 
    col = colorRamp2(c(-2, 0, 2), c("green", "white", "red")),
    column_title = "dist")
Heatmap(mat_with_outliers, name = "mat", 
    col = colorRamp2(c(-2, 0, 2), c("green", "white", "red")),
    clustering_distance_rows = robust_dist,
    clustering_distance_columns = robust_dist,
    column_title = "robust_dist")
```
## Chapter 1.3.2 Render dendrograms

```{r}
library(dendextend)
row_dend = as.dendrogram(hclust(dist(mat)))
row_dend = color_branches(row_dend, k = 2) # `color_branches()` returns a dendrogram object
Heatmap(mat, name = "mat", cluster_rows = row_dend)

library(dendextend)
row_dend = as.dendrogram(hclust(dist(mat)))
row_dend = color_branches(row_dend, k = 5) # `color_branches()` returns a dendrogram object
Heatmap(mat, name = "mat", cluster_rows = row_dend)
```

```{r}
row_dend = dendrapply(row_dend, function(d) {
    attr(d, "nodePar") = list(cex = 0.8, pch = sample(20, 1), col = rand_color(1))
    return(d)
})
Heatmap(mat, name = "mat", cluster_rows = row_dend, row_dend_width = unit(2, "cm"))

row_dend = dendrapply(row_dend, function(d) {
    attr(d, "nodePar") = list(cex = 0.8, pch = sample(20, 1), col = rand_color(2))
    return(d)
})
Heatmap(mat, name = "mat", cluster_rows = row_dend, row_dend_width = unit(4, "cm"))

#The pch in R defines the point symbols in the functions plot() and lines().\
    #https://r-lang.com/pch-in-r/
# cex: number indicating the amount by which plotting text and symbols should be scaled relative to the default. 1=default, 1.5 is 50% larger, 0.5 is 50% smaller, etc. 
    #https://www.statmethods.net/advgraphs/parameters.html

```
## Chapter 1.3.3 Reordering Row Dendogram
```{r}
m2 = matrix(1:100, nr = 10, byrow = TRUE)
Heatmap(m2, name = "mat", row_dend_reorder = FALSE, column_title = "no reordering")
Heatmap(m2, name = "mat", row_dend_reorder = TRUE, column_title = "apply reordering")
```
## Chapter 1.3.4 Reordering Col Dendogram
```{r}
Heatmap(mat, name = "mat", column_title = "default reordering")

library(dendsort)
col_dend = dendsort(hclust(dist(t(mat))))
Heatmap(mat, name = "mat", cluster_columns = col_dend,
    column_title = "reorder by dendsort")
```
## Chapter 1.4 Set row and column orders
```{r}
Heatmap(mat, name = "mat", row_order = order(as.numeric(gsub("row", "", rownames(mat)))), 
    column_order = order(as.numeric(gsub("column", "", colnames(mat)))),
    column_title = "reorder matrix")

Heatmap(mat, name = "mat", row_order = sort(rownames(mat)), 
    column_order = sort(colnames(mat)),
    column_title = "reorder matrix by row/column names")
```


## Chapter 1.5 Set dimension names

```{r}
Heatmap(mat)
Heatmap(mat, name = "mat", show_row_names = FALSE, show_column_names = FALSE)
```


```{r}
Heatmap(mat)
Heatmap(mat, name = "mat", row_names_side = "left", row_dend_side = "right", 
    column_names_side = "top", column_dend_side = "bottom")
```


```{r}
Heatmap(mat, name = "mat", column_names_rot = 0, column_names_side = "top",
    column_dend_side = "bottom")
Heatmap(mat, name = "mat", column_names_rot = 45, column_names_side = "top",
    column_dend_side = "bottom")
```

## Chapter 1.7  Heatmap split

```{r}
#Split based on densogram

Heatmap(mat, name = "mat")
Heatmap(mat, name = "mat", row_split = 2, column_split = 3)
```


```{r}
#SPlit based on Kmeans clusterinh
Heatmap(mat, name = "mat")
Heatmap(mat, name = "mat", row_km = 5)
Heatmap(mat, name = "mat", column_km = 5)
```

```{r}
# Custom split based on external parameters, using a list 

rowsplit <- rep(c("A", "B"), 9)

Heatmap(mat, name = "mat", 
    row_split = rowsplit)

#You can roorder the split by reordering alevel of the factor 
```

```{r}
# Custom split based on external parameters, split by a data frame
splitdf <- data.frame(rep(c("A", "B"), 9), rep(c("C", "D"), each = 9))

Heatmap(mat, name = "mat", 
    row_split = splitdf)

Heatmap(mat, name = "mat", 
    row_split = splitdf,
    row_gap = unit(5, "mm"))
```


```{r}
#More customization of the Tittles 

Heatmap(mat, name = "mat", 
    column_km = 3, column_title_gp = gpar(fill = c("red", "blue", "green"), font = 1:3),
    column_names_gp = gpar(col = c("green", "orange", "purple"), fontsize = c(10, 14, 8)))

```

